---
title: "Fast gene set enrichment with R"
author: Michael Jahn
date: 2023-07-31
output:
  html_notebook:
    theme: cosmo
    toc: no
    number_sections: no
  html_document:
    toc: no
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fast gene set enrichment

### Background

In contrast to a simple gene set enrichment with the hypergeometric test, fast gene set enrichment (**`fgsea`**) does not only test for over- or under-representation of items (groups) in a defined global set, but also takes a quantitative score (expression strength, fold change, etc) into account.

This document walks through one example where the the Bioconductor package `fgsea` is used to determine gene enrichment for pathways of the [subtiwiki pathway categories](http://www.subtiwiki.uni-goettingen.de/v3/category/ for the bacterium *Bacillus subtilis*.

### Libraries and test data

- Install `fgsea` package from Bioconductor

```{r, eval = FALSE}
BiocManager::install("fgsea")
```


- load required libraries

```{r}
suppressPackageStartupMessages({
  library(fgsea)
  library(tidyverse)
})
```

- import category test data (from subtiwiki)

```{r}
df_subtiwiki <- read_csv("../data/geneCategories-2022-09-02.csv",
  name_repair = function(x) str_replace_all(x, " ", "_"),
  show_col_types = FALSE)

head(df_subtiwiki)
```

- `fgsea` requires a gene expression matrix and gene-pathway annotation as input
- can calculate arbitrarily small p-values using iterative random sampling
- bring input into right format

```{r}
df_pathway <- df_subtiwiki %>%
  select(-gene_name, -category_id) %>%
  distinct() %>%
  rename(locus_tag = gene_locus) %>%
  mutate(locus_tag = gsub("_", "", locus_tag))

df_expression <- df_pathway %>%
  select(locus_tag) %>%
  distinct() %>%
  mutate(condition = "test", log2FC = rnorm(n(), 0, sd = 3))

list_fgsea_pathways <- df_pathway %>%
    group_by(category) %>%
    summarize(locus_tag = list(locus_tag)) %>%
    deframe

head(list_fgsea_pathways, n = 3)
```

### Generalized function to run `fgsea` 

- we can make a generalized function that can run `fgsea` on arbitrary data
- can run on many categories at once
- takes as input:
  - `gene_table`: data frame with columns `condition`, `log2FC`, `locus_tag`
  - `conditions`: optional vector of conditions where tests are run independently; requires a matching column in `gene_table` named `condition`
  - `pathways`: named list of pathways containing vectors of `locus_tag`s
  - `collapse`: logical that defines whether similar pathways should be collapsed
  

```{r}
fgsea_multiple <- function(
    conditions, gene_table, pathways,
    minSize = 5, maxSize = 200, collapse = TRUE) {
  result <- lapply(conditions,
    function(cond){
      genes <- gene_table %>%
        filter(condition == cond, !is.na(log2FC), !is.na(locus_tag)) %>%
        select(locus_tag, log2FC) %>%
        deframe
      df_fgsea <- fgsea(
        pathways, genes,
        minSize = minSize, maxSize = maxSize
      )
      if (collapse) {
          main_pws <- collapsePathways(
          df_fgsea,
          pathways,
          genes
        )$mainPathways
        df_fgsea <- filter(df_fgsea, pathway %in% main_pws)
      }
      return(df_fgsea)
  })
  names(result) <- conditions
  return(bind_rows(result, .id = "condition"))
}
```


### Run test and inspect results

- set a seed to obtain same result every time

```{r}
set.seed(123)

fgsea_result <- fgsea_multiple(
  conditions = "test",
  gene_table = df_expression,
  pathways = list_fgsea_pathways
)
```

- expected result: adjusted p-values should not be significant (no enrichment of a specific pathway)

```{r}
fgsea_result %>%
  arrange(padj) %>%
  head(n = 5)
```

- what happens if one pathway has higher log2 FC?
- it suddenly becomes highly significant

```{r}
df_expression <- df_expression %>%
  mutate(log2FC = case_when(
    locus_tag %in% list_fgsea_pathways[["Biosynthesis of peptidoglycan"]] ~ rnorm(n()) - 6,
    .default = log2FC
  ))
```


```{r}
set.seed(123)

fgsea_result <- fgsea_multiple(
  conditions = "test",
  gene_table = df_expression,
  pathways = list_fgsea_pathways
)

fgsea_result %>%
  arrange(padj) %>%
  head(n = 5)
```

