---
title: "Retrieve KEGG Brite pathway information"
author: Michael Jahn
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    theme: cosmo
    toc: no
    number_sections: no
  html_document:
    toc: no
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## KEGG (Brite) pathways

### Background

- Uniprot does not have good pathway annotation
- GO terms (obtained from Uniprot) are sometimes difficult to work with as they are too specific or unspecific
- GO terms have the (dis-) advantage of labelling a gene with many different functional classes
- what we often like to have instead is a simple overview about gene-pathway relationship
- simple means here that one gene is associated with only one or few descriptive functions
- KEGG (Brite) is a manually curated (not homology based) database that offers such information
- this example uses the Bioconductor package `KEGGRest` to retrieve pathways for the bacterium *Bacillus subtilis*.

### Libraries and test data

- Install `KEGGREST` package from Bioconductor

```{r, eval = FALSE}
BiocManager::install("KEGGREST")
```


- load required libraries

```{r}
suppressPackageStartupMessages({
  library(KEGGREST)
  library(tidyverse)
})
```


### Retrieve pathways

#### Starting with organism ID

- KEGG uses organism IDs, for *Bacillus subtilis* it is `bsu`
- using this ID, we can retrieve gene-pathway relationships using a premade R function
- internally it uses the `keggLink` function to find pathways and `keggList` to retrieve human readable pathway names
- it also trims some unnecessary text
- it can be used with organism ID (example: `bsu`) or gene ID (example: `bsu:BSU00040`)

```{r}
source("../source/get_kegg_pathways.R")
```

- apply function to retrieve pathways by organism

```{r}
df_kegg <- get_kegg_pathways(id = "bsu")
head(df_kegg)
```

#### Starting with gene ID

- we can supply one or more IDs but they need to have the organism tag in the front

```{r}
df_kegg_genes <- get_kegg_pathways(id = c("bsu:BSU00040", "bsu:BSU20340"))
head(df_kegg_genes)
```

### Results

- overview about most abundant pathways by locus_tag
- only top 10 pathways are shown

```{r}
df_summary <- df_kegg %>%
  group_by(kegg_pathway) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:10)

head(df_summary)
```

```{r, fig.width = 6.5, fig.height = 3.5}
df_summary %>%
  slice(10:1) %>%
  ggplot(aes(x = count, y = fct_inorder(kegg_pathway))) +
  geom_col()
```

- how many genes are associated with multiple pathways?
- this anaylsis is the inverse of the previous one
- 524 genes are associated with only 1 pathway, 241 with 2, and so on

```{r}
df_summary <- df_kegg %>%
  group_by(locus_tag) %>%
  summarize(genes_per_pathway = n()) %>%
  count(genes_per_pathway)

df_summary
```

```{r, fig.width = 6.5, fig.height = 3.5}
df_summary %>%
  ggplot(aes(x = genes_per_pathway, y = n)) +
  geom_line() +
  geom_point()
```

