---
title: "Hypergeometric test in R"
author: Michael Jahn
date: 2022-01-11
output:
  html_notebook:
    theme: cosmo
    toc: no
    number_sections: no
  html_document:
    toc: no
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hypergeometric test

### Background

The hypergeometric test is a simple statistical procedure to test for over- or under-representation of items (groups) in a defined global set. For example, if a global set consists of 50 black and 50 white balls, and 20 balls are randomly selected from this set, what is the probability that 17 white balls and only 3 black balls are chosen? The hypergeometric test can answer such questions by using the probability function of the [hypergeometric distribution](https://en.wikipedia.org/wiki/Hypergeometric_distribution). This distribution is highly related to the binomial distribution which uses a similar definition but *with replacement*, instead of *without*.

This document walks through one example where the hypergeometric test is used to determine gene enrichment for pathways of the [subtiwiki pathway categories](http://www.subtiwiki.uni-goettingen.de/v3/category/ for the bacterium *Bacillus subtilis*.

### Libraries and test data

- Install `Category` package from Bioconductor

```{r, eval = FALSE}
BiocManager::install("Category")
```


- load required libraries

```{r}
suppressPackageStartupMessages({
  library(Category)
  library(tidyverse)
})
```

- import category test data (from subtiwiki)

```{r}
df_subtiwiki <- read_csv("../data/geneCategories-2022-09-02.csv",
  name_repair = function(x) str_replace_all(x, " ", "_"),
  show_col_types = FALSE)

head(df_subtiwiki)
```

- convert dataframe into a nested list format

```{r}
subti_categories <- df_subtiwiki %>%
  mutate(category = str_sub(category, 1, 25)) %>%
  select(category, gene_locus) %>%
  group_by(category) %>%
  summarize(gene_locus = list(gene_locus)) %>%
  deframe()

head(subti_categories, n = 3)
```

### Prepare input data

- **assayed**: gene sets that were tested in an experiment, usually all genes/pathways in the universe
- example: we select a subset of 20 subtiwiki categories with each more than 20 genes

```{r}
assayed <- subti_categories[lapply(subti_categories, length) >= 20][1:20]
```

- **significant**: genes within the assayed set that were observed, or fall below a sign. threshold
- example: we select 10% of genes as significant

```{r}
significant <- lapply(assayed, function(x) sample(x, length(x)/10))
```

- **universe**: all unique gene names

```{r}
universe = assayed %>%
  unlist(use.names = FALSE) %>%
  unique
```

### Run test and inspect results

- run hypergeometric test using `hyperg` function from `Categories`

```{r}
result <- hyperg(
  assayed,
  significant,
  universe,
  representation = "over"
)

result <- mutate(result, across(everything(), unlist))
```

- expected result: all p-values are very similar (no enrichment of a specific pathway)

```{r}
head(result)
```

- what happens if one pathway has more than 10% sign. genes?

```{r}
significant[[1]] <- sample(assayed[[1]], 20)
```

- run hypergeometric test again and inspect result
- the first (modified) pathway has lower p-value and higher odds

```{r}
hyperg(
    assayed,
    significant,
    universe,
    representation = "over"
  ) %>% 
  mutate(across(everything(), unlist)) %>%
  head
```

