---
title: "Quantifiying overlap in the composition of categories"
author: Michael Jahn
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    theme: cosmo
    toc: no
    number_sections: no
  html_document:
    toc: no
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

Often in biology we encounter situations where one gene maps to different pathways, or different pathways to the same gene.
I other words, gene-pathway relationships are member-category relationships that can have a large degree of redundancy or overlap. Some pathways/categories might be highly similar when comparing the list of their members. It is often useful to know the degree of overlap between pathways/categories, and optionally reduce the list of pathways to leave only one representative instead of multiple highly similar ones.

This document walks through one example where we determine overlap between pathways of the [subtiwiki pathway categories](http://www.subtiwiki.uni-goettingen.de/v3/category/ for the bacterium *Bacillus subtilis*.

If you find a built-in function in an R package that will do this automatically, please [leave a message](https://github.com/MPUSP/bioinfo-code-chunks/issues).

### Libraries and test data

- load required libraries

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

- import category test data (from subtiwiki)

```{r}
df_subtiwiki <- read_csv("../data/geneCategories-2022-09-02.csv",
  name_repair = function(x) str_replace_all(x, " ", "_"),
  show_col_types = FALSE)

head(df_subtiwiki)
```

### Calculate overlap

- in order to estimate overlap, the data is transformed into a 2-dimensional matrix with pathways on rows, genes on columns
- the matrix is a binary matrix filled with membership relations (1 - member, 0 - no member)

```{r}
mat_categories <- df_subtiwiki %>%
  mutate(category = str_sub(category, 1, 25)) %>%
  select(category, gene_locus) %>%
  distinct() %>% 
  mutate(dummy = 1) %>%
  pivot_wider(id_cols = category, names_from = gene_locus,
    values_from = dummy, values_fill = 0) %>%
  column_to_rownames("category") %>%
  as.matrix

mat_categories[1:5, 1:5]
```

- reduce result to 50 first categories
- the binary matrix  is clustered by similarity and plotted as dendrogram

```{r}
hclust_cat <- dist(mat_categories[1:50, ], method = "binary") %>%
  hclust(method = "complete")

plot(hclust_cat, cex = 0.6, main = NA)
```


- the binary distance metric shows that, for example, two pathways with distance 0.5 have 50% of their member genes in common
- we can find these by "cutting" the dendrogram at height = 0.5

```{r}
cutree(hclust_cat, h = 0.5) %>%
  {duplicated(.) | rev(duplicated(rev(.)))} %>%
  which %>% hclust_cat$labels[.]
```

- we can also calculate the distance metric directly and find similar pathways by filtering
- note: the heatmap shows only every second label due to size restrictions

```{r, fig.width = 5.5, fig.height = 5.5}
dist(mat_categories[1:50, ], method = "binary") %>%
  as.matrix %>%
  heatmap
```
- filter categories (rows) where dist metric falls below a threshold of 0.5
- same result as dendrogram

```{r}
dist(mat_categories[1:50, ], method = "binary") %>%
  as.matrix %>%
  as.data.frame %>%
  mutate(across(everything(), ~ replace(.x, .x == 0, NA))) %>%
  filter(if_any(everything(), ~ .x < 0.5))
```


